\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{geometry}
\geometry{a4paper, margin=1in}

\title{Unifying Sequence-to-Function Models with Continuous Flow Matching with STREAM}
\author{}
\date{}

\begin{document}

\maketitle

We propose Sequence-conditioned Trajectories via Regulatory Element Matching (STREAM), a framework for learning continuous gene expression dynamics from time-resolved single-cell data. STREAM unifies sequence-to-function models with continuous flow matching, enabling the inference of gene regulatory dynamics across developmental time.

\section{Background}

By training to predict diverse chromatin marks, sequence-to-function models learn representations of regulatory DNA sequences, in particular candidate cis-regulatory elements (cCRE). The representation ideally summarizes sequence features relevant to gene regulation, such as transcription factor motifs and higher-order sequence grammar. 

The state-of-the-art paradigm for learning ordinary differential equation (ODE) models parameterized by a continuous-time vector field is Conditional Flow Matching (CFM). The CFM ``trick'' is framing model training as a regression that fits target velocities along simple paths between data distributions at different time points. This avoids the high computational cost of forward simulating the ODE, enabling efficient, scalable training. For time-resolved single-cell data, this corresponds to fitting a vector field that transports cell-state distributions between observed time points.

\section{Model setup}

Continuous cell state is represented as $x_t \in \mathbb{R}^G$ at time $t$, with $G$ being the number of genes in the selected modeling panel. For the mouse development implementation, $G$ is a configurable protein-coding highly variable gene subset so that STREAM and the baseline CFM are compared on exactly the same output coordinates. We observe the expression profiles of cells at a discrete set of time points (e.g., embryonic days) $ \mathcal{T} = \{t_0, t_1, \dots, t_K\} $. 
To make training comparable across datasets we require $t$ is on the same scale (e.g. days). The empirical data distribution at time $t_k$ is $p_k(x)$.

Let $\mathcal{C}_g$ be the set of cCREs associated with gene $g$. In the mouse development implementation, cCREs are linked to genes if their midpoint is within 100 kb of the gene transcription start site (TSS) in mm10. Let $z_c \in \mathbb{R}^d$ denote a precomputed representation of cCRE $c$, generated by AlphaGenome from local DNA sequence. The set of all precomputed cCRE embeddings for gene $g$ is $\mathbf{Z}_g = \{ z_c \mid c \in \mathcal{C}_g \}$.

For each gene, one token is distinguished as the promoter token. The promoter token is the linked cCRE closest to the TSS if that cCRE is within 1 kb of the TSS. If the closest linked cCRE is farther than 1 kb, a synthetic promoter cCRE centered on the TSS is inserted. The promoter token is placed first in the token sequence. This guarantees that every gene has an explicit promoter representation even if distal cCRE annotation is sparse near the TSS.

We establish a sequence and state dependent map $f$ from each gene's sequence-derived regulatory elements and the instantaneous global cell state to that gene's velocity. The cCREs associated with a gene are treated as a sentence of tokens processed by a transformer. Token features include the AlphaGenome embedding, signed distance to TSS, a promoter indicator, and optional positional information. The implementation supports no positional encoding, learned binned distance embeddings, and rotary positional embeddings (RoPE) applied using signed TSS-relative positions. Cell state $x_t$ conditions the regulatory token processing either through FiLM or through cross-attention. The scalar output for each gene is read from the final promoter-token representation, analogous to a CLS readout.

We define a global vector field,
$$ \frac{dx_t}{dt} = \begin{bmatrix} f(x_t, \mathbf{Z}_1) \\ \vdots \\ f(x_t, \mathbf{Z}_G) \end{bmatrix} \in \mathbb{R}^G $$
This allows the model to capture \emph{trans}-regulatory effects because each gene-specific velocity depends on the full cell state $x_t$, not only on that gene's expression level.

Note that we do not condition on time $t$ to allow generalization across different time-scales. Additionally, genes are represented as a set of regulatory elements, rather than a single token, to force the model to learn global regulatory grammar rather than memorizing gene-specific dynamics.

\paragraph{Conditioning variants.} The FiLM STREAM variant maps $x_t$ to feature-wise scale and shift parameters that modulate regulatory token states before promoter readout. The cross-attention STREAM variant maps $x_t$ to a small set of context tokens; regulatory tokens attend to these context tokens before promoter readout. Both variants share the same cCRE-gene links, AlphaGenome embeddings, promoter-token convention, minibatch OT coupling, and selected gene panel.

\subsection{Architecture of the gene velocity model}

For each selected gene $g$, the implementation packs up to $M$ regulatory tokens into a fixed tensor. Token $m$ for gene $g$ has an AlphaGenome embedding $z_{gm} \in \mathbb{R}^{3072}$, signed TSS-relative distance $d_{gm}$, a promoter indicator $p_{gm}$, and a mask $a_{gm}$ indicating whether the token is real rather than padding. The promoter token is always placed at index $m=0$. Let
$$
e_{gm}^{(0)} =
W_z z_{gm} +
\phi_d(d_{gm}/100{,}000) +
E_p[p_{gm}] +
\phi_{\mathrm{pos}}(d_{gm}),
$$
where $W_z$ projects AlphaGenome embeddings to the model width, $\phi_d$ is a small MLP for signed distance, $E_p$ is a learned promoter/non-promoter embedding, and $\phi_{\mathrm{pos}}$ is optional positional information. In the current code, $\phi_{\mathrm{pos}}$ can be absent, a learned binned-distance embedding, or RoPE applied to token features using signed distance. Padding tokens are masked in all self-attention layers.

The gene regulatory sequence is encoded by a shared transformer whose layers are interleaved with cell-state conditioning,
$$
H_{bg:}^{(\ell + 1)} =
\mathrm{Cond}^{(\ell)}
\left(
\mathrm{TransformerLayer}^{(\ell)}(H_{bg:}^{(\ell)}), x_{bt}
\right),
$$
where $H_{g:}^{(0)} = (e_{g1}^{(0)}, \ldots, e_{gM}^{(0)})$ is broadcast across cells in the minibatch and $\ell$ indexes transformer layers. The same layer parameters are used for every gene. This weight sharing is important: the model must learn regulatory grammar that transfers across genes rather than assigning each gene a free velocity parameter. The default mouse development configuration uses model width 256, 8 attention heads, 4 transformer layers, dropout 0.1, and at most 32 CRE/promoter tokens per gene.

The cell state $x_t$ is incorporated at every transformer layer. This makes $f$ both sequence dependent and state dependent throughout regulatory token processing. Two conditioning mechanisms are implemented:

\paragraph{FiLM conditioning.} At each transformer layer $\ell$, a layer-specific cell-state MLP maps $x_t$ to feature-wise scale and shift vectors,
$$
(\gamma^{(\ell)}(x_t), \beta^{(\ell)}(x_t)) = h_{\mathrm{FiLM}}^{(\ell)}(x_t), \qquad
H_{bgm}^{(\ell + 1)} = U_{bgm}^{(\ell)} \odot (1 + \gamma_b^{(\ell)}) + \beta_b^{(\ell)},
$$
for cell $b$, gene $g$, token $m$, and transformer output $U^{(\ell)}$. A layer normalization and scalar head are then applied only to the final promoter token,
$$
f_g(x_t,\mathbf{Z}_g) = w^\top \mathrm{LN}(H_{bg0}^{(L)}) + b.
$$

\paragraph{Cross-attention conditioning.} At each transformer layer $\ell$, a layer-specific cell-state MLP maps $x_t$ to a small set of context tokens $C_b^{(\ell)} \in \mathbb{R}^{L_c \times D}$, with $L_c=8$ by default. Regulatory tokens query these context tokens after CRE self-attention,
$$
H_{bg:}^{(\ell + 1)} =
U_{bg:}^{(\ell)} +
\mathrm{MHA}^{(\ell)}(Q=U_{bg:}^{(\ell)}, K=C_b^{(\ell)}, V=C_b^{(\ell)}),
$$
and the same promoter-token scalar head gives $f_g(x_t,\mathbf{Z}_g)$. This variant allows each gene's regulatory tokens to extract different information from the global cell state.

The resulting vector field is formed by evaluating the same gene-level module for all selected genes:
$$
f_\theta(x_t, \{\mathbf{Z}_g\}_{g=1}^G)
=
\left(f_1(x_t,\mathbf{Z}_1), \ldots, f_G(x_t,\mathbf{Z}_G)\right).
$$
The implementation does not use a learned gene identity embedding by default. Gene specificity enters through the set of linked CRE/promoter tokens and their sequence-derived embeddings.

\paragraph{Expression-only baseline.} The standard CFM baseline uses the same selected genes, same minibatch OT coupling, same interpolation path, and same target velocity, but replaces the sequence-conditioned gene module with an MLP $f_{\mathrm{CFM}}(x_t): \mathbb{R}^G \rightarrow \mathbb{R}^G$. This isolates the contribution of CRE/AlphaGenome conditioning from the flow matching training setup.

\paragraph{UCE cell-state ablation.} We additionally evaluate a frozen 33-layer Universal Cell Embedding (UCE) representation $u=E(x)\in\mathbb{R}^{1280}$ as the input state to the vector field. OT is still solved using the selected expression panel, and the target velocity remains expression-valued. After sampling an OT pair $(x_k,x_{k+1})$ and a CFM time $\tau$, we interpolate both spaces with the same pair and time:
$$x_t=(1-\tau)x_k+\tau x_{k+1},\qquad u_t=(1-\tau)E(x_k)+\tau E(x_{k+1}).$$
The UCE baseline predicts expression velocity as $f_{\mathrm{CFM}}(u_t):\mathbb{R}^{1280}\rightarrow\mathbb{R}^{G}$; FiLM and cross-attention STREAM use $u_t$ in every layer-specific conditioning module while retaining the same CRE inputs and expression-valued promoter-token outputs. Thus UCE changes the state representation supplied to $f$, not the state space in which the dynamics or loss are defined.

\paragraph{Sequential Flow Matching Objective.} To fit $f$ across multiple time points, we match it to a target marginal vector field $\hat{f}_t$. We define independent flow matching objectives between consecutive time points $[t_k, t_{k+1}]$. Let $x_k \sim p_k(x)$ and $x_{k+1} \sim p_{k+1}(x)$ be sampled from the empirical distributions. We construct a linear path mapping $x_k$ to $x_{k+1}$. For $t \in [t_k, t_{k+1}]$, define the normalized time step $\tau = \frac{t - t_k}{t_{k+1} - t_k}$.

The interpolated state $x_t$ is:
$$ x_t(\tau) = (1 - \tau) x_k + \tau x_{k+1} $$

The target conditional vector field, representing the velocity required to move from $x_k$ to $x_{k+1}$ is,
$$ \hat{f}_t = \frac{x_{k+1} - x_k}{t_{k+1} - t_k} $$

The unified Conditional Flow Matching (CFM) objective minimizes the expected squared difference between the sequence-conditioned network $f$ and the target vector field $\hat{f}_t$ across all consecutive time intervals:
$$ \mathcal{L}(f) = \sum_{k=0}^{K-1} \mathbb{E}_{t \sim \mathcal{U}[t_k, t_{k+1}], \; x_k, x_{k+1} \sim p_{data}(x_k, x_{k+1})} \left[ \left\| f(s_t, \{\mathbf{Z}_g\}_{g=1}^G) - \hat{f}_t \right\|^2 \right] $$
where $s_t=x_t$ for the expression-state experiments and $s_t=u_t$ for the UCE-state ablation.

While it is valid to use the trivial joint distribution $p_{data}(x_k, x_{k+1}) = p_{k}(x_k) p_{k+1}(x_{k+1})$, STREAM uses minibatch optimal transport to couple cells between neighboring time points. Each minibatch contains cells from exactly one interval $(t_k,t_{k+1})$. A pairwise cost is computed between sampled cells at $t_k$ and sampled cells at $t_{k+1}$, an entropic OT coupling is solved on the minibatch, and training pairs are sampled from the resulting coupling.

\section{Mouse development implementation}

The implementation targets the JAX mouse development AnnData files and mm10 references. Gene models are read from the Ensembl GRCm38 GTF, and cCREs are read from the mouse cCRE BED file. TSS coordinates are converted from 1-based GTF coordinates to 0-based genomic positions. For positive-strand genes the TSS is the gene start; for negative-strand genes it is the gene end.

AlphaGenome embeddings are computed once and cached. The implementation uses the local PyTorch AlphaGenome port, mouse organism index 1, and the 128 bp trunk embeddings. For each real or synthetic cCRE window, the 128 bp embeddings are pooled to one fixed vector. Training reads the cached CRE token tensor and does not call AlphaGenome online.

\section{Baseline and evaluation}

STREAM is compared against a standard expression-only CFM baseline. The baseline uses the same selected genes, same neighboring-time minibatch sampler, same minibatch OT coupling, same interpolation path, and same velocity target, but it does not receive cCRE or AlphaGenome inputs. Both baseline and STREAM are run with expression-state and UCE-state inputs, with UCE results written separately from expression-state checkpoints.

Evaluation is performed on held-out time points. Whole embryonic days are removed from training intervals. Training uses only adjacent intervals whose two endpoints are both training time points. Evaluation reports held-out-timepoint losses and rollout diagnostics on intervals adjacent to or involving held-out days when those endpoints are available. This directly tests temporal generalization rather than embryo-level generalization.

\section{Cross-species zebrafish transfer}

We evaluate whether the UCE-conditioned STREAM field learned from mouse
development transfers to zebrafish development. The target data are the
wild-type ZSCAPE reference atlas (GEO GSE202639), restricted to uninjected
28C controls. We use the GRCz11 Ensembl reference and the ZEPA developmental
single-cell chromatin accessibility cCRE catalogue. The optional dynamic-cCRE
supplement is included when its workbook exposes readable genomic coordinates.
Elements within 100 kb of a protein-coding HVG TSS
are linked to the gene, with at most 32 packed tokens and exactly one promoter
token. The 33-layer UCE encoder is frozen and uses zebrafish gene-token
metadata; OT, targets, and outputs remain in expression space.

For each 5k and 10k panel, FiLM and cross-attention STREAM are evaluated as
(i) a frozen mouse checkpoint applied zero-shot to zebrafish, (ii) the same
checkpoint fine-tuned on zebrafish, and (iii) a zebrafish-only initialization.
Standard CFM is trained from scratch on zebrafish using the same genes and
minibatch OT protocol. We hold out 36 and 72 hpf rather than embryos and score
only intervals touching those stages.

We compare physical time, $t=\mathrm{hpf}/24$, with organism-relative time,
where each species' observed development is mapped to $[0,1]$. Separate mouse
source models are trained for each convention. Native CFM loss is reported
within a time convention; a scale-aware expression-displacement metric,
$\lvert (f_\theta-\hat f)(t_{k+1}-t_k) \rvert$, is compared across them.
Held-out endpoints are cached and OT sampling is seeded for every variant.

The published AlphaGenome model exposes only human and mouse organism
conditioning. GRCz11 sequence windows therefore use its mouse organism index,
which is an explicit out-of-distribution assumption rather than zebrafish
calibration.

\end{document}
